---
title: "Bird Data.qmd"
format: html
editor: visual
---

# Bird Data

Bird Survey Data including date, time, surveyors, site, temp, cloud cover, windspeed, survey point, species spotted, and distance

## LOAD AND CLEAN DATA

### Load Libraries

```{r}
library(ggplot2)
library(tidyverse)
library(vegan)
```

### Load Data

```{r}
#Load csv files
bird1 <-read.csv("Bird Survey Data - December 2023.csv")
bird2 <-read.csv("Bird Survey Data - February 2024.csv")
```

```{r}
#Merge data sets
bird <- bind_rows(bird1, bird2) 
#Make a new column that combines the Site and Survey point
bird$SiteID<- paste(bird$Site, bird$Point)
#Change date from mm/dd/yyyy to yyyy-mm-dd
bird$Date<-as.Date(bird$Date, "%m/%d/%y")

```

## DATA SUMMARIES

### Summary of Species Counts

```{r}
#Summarize count of each bird species found at each site
birdcount <- bird %>%
  group_by(SiteID, Species, Date, Site) %>%
  summarise(Count = n()) %>%
  ungroup()

```

### Presence/Absence 

```{r}
#These lines of code add 0 to species that were observed in one siteID but not the others
bird_zero <- expand.grid(Species = unique(birdcount$Species), SiteID = unique(birdcount$SiteID)) 
#Joins bird_zero and birdcount into one dataframe by matching Species and Site ID
bird_zero <- left_join(bird_zero, birdcount, by = c("Species", "SiteID"))
#If bird count is NA, replace with 0
bird_zero$Count[is.na(bird_zero$Count)] <- 0
```

### Log Transformation

```{r}
#log transform data
bird_zero$Countln <- log((bird_zero$Count)+1)
#Only keep specified columns
bird_zero<- bird_zero %>%
  select(Species, SiteID, Count, Countln)
```

# \*\*Abundance

```{r}

```

## SPECIES RICHNESS

### By Site

```{r}
#Species richness at each survey point
species_richness_point<- birdcount %>%
  group_by(SiteID) %>%
  summarize(Richness = n_distinct(Species))

```

#### Graph

```{r}
#Graph species richness by survey point
ggplot() +
  geom_bar(data = species_richness_point, aes(x = SiteID, y = Richness, fill = ifelse(grepl("^BSB", SiteID), "BSB", "CDL")), stat = "identity", position = "dodge", color = "black") +
  labs(title = "Species Richness Between Survey Points",
       x = "Survey Point", y = "Number of Species") +
  scale_fill_manual(values = c("BSB" = "tan1", "CDL" = "deepskyblue4")) +
  theme(plot.title = element_text(hjust = 0.5))+
  scale_y_continuous(expand = c(0, 0),limits = c(0,10)) + 
  theme_classic()+
  guides(fill = FALSE)
```

### By Survey Point

```{r}
#Species richness at each survey site
species_richness_site<- birdcount %>%
  group_by(Site) %>%
  summarize(Richness = n_distinct(Species))
```

#### Graph

```{r}
#Graph species richness by survey site
ggplot() +
  geom_bar(data = species_richness_site, aes(x = Site, y = Richness, fill=Site), stat = "identity", position = "dodge", color = "black") +
  labs(title = "Species Richness Between Survey Points",
       x = "Site", y = "Number of Species") +
  scale_fill_manual(values = c("BSB" = "tan1", "CDL" = "deepskyblue4")) +
  theme(plot.title = element_text(hjust = 0.5))+
  scale_y_continuous(expand = c(0, 0),limits = c(0,20)) + 
  theme_classic()+
  guides(fill = FALSE)
```

## SPECIES DIVERSITY

### Shannon-Diversity Index

```{r}
#Calculate relative abundance of each survey point
birdcount2 <- birdcount %>%
  group_by(SiteID, Date) %>%
  mutate(Relative_Abundance = Count / sum(Count))

#Calculate the Shannon-diversity index for each survey point separated by date
shannon_index <- birdcount2 %>%
  group_by(SiteID, Site, Date) %>%
  summarize(Shannon_Index = -sum(Relative_Abundance * log(Relative_Abundance)))
```

### By Site

```{r}
#Calculate the average Shannon-diversity index for each survey point
surveysite_diversity <- shannon_index %>%
  group_by(Site) %>%
  summarise(across(where(is.numeric), mean, na.rm = TRUE))
```

#### Graph

```{r}
#Graph by total in each survey site
ggplot() +
  geom_bar(data = surveysite_diversity, aes(x = Site, y = Shannon_Index, fill = ifelse(grepl("^BSB", Site), "BSB", "CDL")), stat = "identity", position = "dodge", color = "black") +
  labs(title = "Species Diversity Comparison Between Survey Points",
       x = "Survey Point", y = "Shannon-Diverity Index") +
  scale_fill_manual(values = c("BSB" = "tan1", "CDL" = "deepskyblue4")) +
  theme(plot.title = element_text(hjust = 0.5))+
  scale_y_continuous(expand = c(0, 0),limits = c(0,1.5)) + 
  theme_classic()+
  guides(fill = FALSE)
```

### By Survey Point

```{r}
#Calculate the average Shannon-diversity index for each survey point
surveypoint_diversity <- shannon_index %>%
  group_by(SiteID) %>%
  summarise(across(where(is.numeric), mean, na.rm = TRUE))
```

#### Graph

```{r}
#Graph by total in each survey point
ggplot() +
  geom_bar(data = surveypoint_diversity, aes(x = SiteID, y = Shannon_Index, fill = ifelse(grepl("^BSB", SiteID), "BSB", "CDL")), stat = "identity", position = "dodge", color = "black") +
  labs(title = "Species Diversity Comparison Between Survey Points",
       x = "Survey Point", y = "Shannon-Diverity Index") +
  scale_fill_manual(values = c("BSB" = "tan1", "CDL" = "deepskyblue4")) +
  theme(plot.title = element_text(hjust = 0.5))+
  scale_y_continuous(expand = c(0, 0),limits = c(0,2)) + 
  theme_classic()+
  guides(fill = FALSE)

```

## ABUNDANCE

### Per Site

```{r}
#total bird count per site
birdtotal <- bird %>%
  group_by(Site) %>%
  summarise(Count = n()) %>%
  ungroup()
```

#### Graph

```{r}
#Graph of abundance at each site
ggplot(birdtotal, aes(x = Site, y = Count, fill=Site)) +
  geom_bar(stat = "identity", color="black") +
  labs(x = "Site", y = "Abundance", title = "Species Abundance by Site") +
  scale_fill_manual(values = c("BSB" = "tan1", "CDL" = "deepskyblue4"))+
  scale_y_continuous(expand = c(0, 0),limits = c(0,35)) +
  theme_classic()+
  theme(plot.title = element_text(hjust = 0.5))
```

### Per Survey Point

```{r}
#total bird count per survey point
birdtotal_site <- bird %>%
  group_by(SiteID) %>%
  summarise(Count = n()) %>%
  ungroup()
```

#### Graph

```{r}
ggplot(birdtotal_site, aes(x = SiteID, y = Count, fill = ifelse(grepl("^BSB", SiteID), "BSB", "CDL"))) +
  geom_bar(stat = "identity", color="black") +
  labs(x = "Site", y = "Abundance", title = "Species Abundance by Site") +
  scale_fill_manual(values = c("BSB" = "tan1", "CDL" = "deepskyblue4"))+
  scale_y_continuous(expand = c(0, 0),limits = c(0,12)) +
  theme_classic()+
  theme(plot.title = element_text(hjust = 0.5))+
  guides(fill = FALSE)
```

### Per Site and Date

```{r}
#total bird count by site and date
birdtotal_dateSite <- bird %>%
  group_by(Site, Date) %>%
  summarise(Count = n()) %>%
  ungroup()
```

#### !Graph

```{r}

```

### Per Survey Point and Date

```{r}
#total bird count by survey point and date
birdtotal_date <- bird %>%
  group_by(SiteID, Date) %>%
  summarise(Count = n()) %>%
  ungroup()
```

#### !Graph

```{r}

```

### Migratory vs Resident

```{r}
migratory <- bird %>%
  group_by(Resident, Site) %>%
  summarise(Count = n())

```

#### Graph

```{r}
ggplot(migratory, aes(x = Resident, y = Count, fill=Site)) +
  geom_bar(stat = "identity", position="dodge", color="black") +
  labs(x="Residency Status", y = "Abundance", 
       title = "Resident vs Migratory Birds by Site") +
  scale_fill_manual(values = c("BSB" = "tan1", "CDL" = "deepskyblue4"))+
  scale_x_discrete(labels = c("No" = "Migratory", "Yes" = "Resident"))+
  scale_y_continuous(expand = c(0, 0), limits = c(0, 25), 
                     labels = scales::number_format(scale = 1, accuracy = 1)) +
  theme_classic()+
  theme(plot.title = element_text(hjust = 0.5))
```

## **\*\*Bird RAI**

Calculate RAI

RAI = \# occurrences/# days in operation

```{r}
birdRAI <- birdcount |> 
  mutate(RAI = Count/2)
 #mutate(MonthYear = mdy(paste(Month,"1",Year,sep = "/"))) 
```

```{r}
#Not the best to represent RAI
birdRAI |> 
  ggplot(aes(y = RAI, x = Date, color = Site)) + 
  geom_point() + 
  geom_line() + 
  facet_wrap(Site ~. )
```
