---
title: "Mesquite Analysis"
format: html
editor: visual
---

# **\*\*IRGA DATA**

### Load Libraries

```{r}
library(tidyverse)
library(lattice)
library(car)
library(scales)
```

## DATA CLEANUP

```{r}
#load and select data
data.irga <- read.csv("GDE_IRGA.csv")

str(data.irga)

data.irga$Site <- as.factor(data.irga$Site)
data.irga$TreeID <- as.factor(data.irga$TreeID)

data.irga <- data.irga %>% subset(Ci > 0)

data.irga.meta <- data.irga %>%
  select(1:2) %>%
  group_by(TreeID) %>%
  unique() %>%
  ungroup()
```

### Calculating Means

```{r}
#CALCULATE AVG PHOTOSYNTHETIC RATE AND STOMATAL CONDUCTANCE FOR EACH TREE AND JOIN GAS EXCHANGE WITH METADATA
data.irga.indiv <- data.irga %>% group_by(TreeID) %>% 
  summarise(Photo = mean(Photo), 
            Cond = mean(Cond),
            WUE = mean(Photo/Cond),
            Count = n()) %>% 
  ungroup() %>%
  merge(data.irga.meta,
        by = "TreeID",
        all.x = TRUE)
```

### Log Transformations

```{r}
#LOG TRANSFORMATIONS
data.irga.indiv$lnPhoto <- log(data.irga.indiv$Photo)
data.irga.indiv$lnCond <- log(data.irga.indiv$Cond)
data.irga.indiv$lnWUE <- log(data.irga.indiv$WUE)
```

### Check Normality

```{r}
#HISTOGRAM: RAW DATA VS LOG DATA
histogram(data.irga.indiv$Photo)
histogram(data.irga.indiv$lnPhoto)

histogram(data.irga.indiv$Cond)
histogram(data.irga.indiv$lnCond)

histogram(data.irga.indiv$WUE)
histogram(data.irga.indiv$lnWUE)
```

### Calculating Standard Error

```{r}
#CALCULATE MEAN AND STANDARD ERROR
data.irga.sum <- data.irga.indiv %>% group_by (Site) %>% 
  summarise(meanPhoto = mean(Photo), 
            sePhoto = sd(Photo)/sqrt(length(Photo)), 
            meanCond = mean(Cond), 
            seCond = sd(Cond)/sqrt(length(Cond)), 
            meanWUE = mean(WUE), 
            seWUE = sd(WUE)/sqrt(length(WUE)), 
  ) %>% ungroup()
```

## GRAPHS

### Photosynthetic Rate

```{r}
#BARPLOT
ggplot(data.irga.sum) + aes(x=Site, y=meanPhoto, fill=Site )+
  scale_fill_manual(values=c("tan1","deepskyblue4"), 
                    labels=c('Borrego Basin', 'Clark Lake')) +
  geom_col(alpha=1, position="dodge") + 
  geom_errorbar( aes(x=Site, ymin=meanPhoto-sePhoto, ymax=meanPhoto+sePhoto), 
                 width=0.4,
                 colour="black", 
                 alpha=0.7, 
                 size=1.0,
                 position=position_dodge(width=0.9)) +
  scale_y_continuous(expand = c(0, 0),limits = c(0,26)) +
  scale_x_discrete(labels=c("1" = "Borrego Basin", "5" = "Clark Lake"))+
  xlab("Site")+ylab("Mean Photosynthetic Rate")+
  ggtitle("Photosynthetic Rates at Each Survey Site")+
  theme_classic()+
  theme(plot.title = element_text(hjust = 0.5))

#BOXPLOT
ggplot(data.irga.indiv) + aes(x=Site, y=Photo, fill=Site)+
  scale_fill_manual(values=c("tan1","deepskyblue4"),
                    labels=c('Borrego Basin', 'Clark Lake')) +
  geom_boxplot(aes(x=Site, y=Photo, fill=Site)) + 
  scale_x_discrete(labels=c("1" = "Borrego Basin", "5" = "Clark Lake"))+
  xlab("Site")+ylab("Photosynthetic Rate")+
  ggtitle("Photosynthetic Rates at Each Survey Site")+
  theme_classic()+
  theme(plot.title = element_text(hjust = 0.5))
```

### Stomatal Conductance

```{r}
#BARPLOT
ggplot(data.irga.sum) + aes(x=Site, y=meanCond, fill=Site )+
  scale_fill_manual(values=c("tan1","deepskyblue4"), 
                    labels=c('Borrego Basin', 'Clark Lake')) +
  geom_col(alpha=1, position="dodge") + 
  geom_errorbar( aes(x=Site, ymin=meanCond-seCond, ymax=meanCond+seCond), 
                 width=0.4,
                 colour="black", 
                 alpha=0.7, 
                 size=1.0,
                 position=position_dodge(width=0.9)) +
  scale_y_continuous(expand = c(0, 0),limits = c(0,0.5)) +
  scale_x_discrete(labels=c("1" = "Borrego Basin", "5" = "Clark Lake"))+
  xlab("Site")+ylab("Mean Stomatal Conductance Rate")+
  ggtitle("Stomatal Conductance Rates at Each Survey Site")+
  theme_classic()+
  theme(plot.title = element_text(hjust = 0.5))

#BOXPLOT
ggplot(data.irga.indiv) + aes(x=Site, y=Cond, fill=Site)+
  scale_fill_manual(values=c("tan1","deepskyblue4"),
                    labels=c('Borrego Basin', 'Clark Lake')) +
  geom_boxplot(aes(x=Site, y=Cond, fill=Site)) + 
  scale_x_discrete(labels=c("1" = "Borrego Basin", "5" = "Clark Lake"))+
  xlab("Site")+ylab("Stomatal Conductance Rate")+
  ggtitle("Stomatal Conductance Rates at Each Survey Site")+
  theme_classic()+
  theme(plot.title = element_text(hjust = 0.5))
```

### Water Use Efficiency

```{r}
#BARPLOT
ggplot(data.irga.sum) + aes(x=Site, y=meanWUE, fill=Site )+
  scale_fill_manual(values=c("tan1","deepskyblue4"), 
                    labels=c('Borrego Basin', 'Clark Lake')) +
  geom_col(alpha=1, position="dodge") + 
  geom_errorbar( aes(x=Site, ymin=meanWUE-seWUE, ymax=meanWUE+seWUE), 
                 width=0.4,
                 colour="black", 
                 alpha=0.7, 
                 size=1.0,
                 position=position_dodge(width=0.9)) +
  scale_y_continuous(expand = c(0, 0),limits = c(0,125)) +
  scale_x_discrete(labels=c("1" = "Borrego Basin", "5" = "Clark Lake"))+
  xlab("Site")+ylab("Mean Stomatal Conductance")+
  ggtitle("Stomatal Conductance Rates at Each Survey Site")+
  theme_classic()+
  theme(plot.title = element_text(hjust = 0.5))

#BOXPLOT
ggplot(data.irga.indiv) + aes(x=Site, y=WUE, fill=Site)+
  scale_fill_manual(values=c("tan1","deepskyblue4"),
                    labels=c('Borrego Basin', 'Clark Lake')) +
  geom_boxplot(aes(x=Site, y=WUE, fill=Site)) + 
  scale_x_discrete(labels=c("1" = "Borrego Basin", "5" = "Clark Lake"))+
  xlab("Site")+ylab("Water Use Efficiency Rate")+
  ggtitle("Water Use Efficiency Rates at Each Survey Site")+
  theme_classic()+
  theme(plot.title = element_text(hjust = 0.5))
```

## STATISTICAL ANALYSIS

### Photosynthetic Rate

```{r}
#ANOVA
Photo.aov <- aov(lnPhoto ~ Site, data=data.irga.indiv)
summary(Photo.aov)

#HOMOGENEITY OF VARIANCES
plot(Photo.aov,1)
leveneTest(lnPhoto ~ Site, data=data.irga.indiv)

#T-TEST
t.test(lnPhoto~Site, data=data.irga.indiv)

#DISTRIBUTION OF RESIDUALS
plot(Photo.aov,2)
```

### Stomatal Conductance

```{r}
#ANOVA
Cond.aov <- aov(lnCond ~ Site, data=data.irga.indiv)
summary(Cond.aov)

#HOMOGENEITY OF VARIANCES
plot(Cond.aov,1)
leveneTest(lnCond ~ Site, data=data.irga.indiv)

#T-TEST
t.test(lnCond~Site, data=data.irga.indiv)

#DISTRIBUTION OF RESIDUALS
plot(Cond.aov,2)
```

### Water Use Efficiency

```{r}
#ANOVA
WUE.aov <- aov(lnWUE ~ Site, data=data.irga.indiv)
summary(WUE.aov)

#HOMOGENEITY OF VARIANCES
plot(WUE.aov,1)
leveneTest(lnWUE ~ Site, data=data.irga.indiv)

#T-TEST
t.test(lnWUE~Site, data=data.irga.indiv)

#DISTRIBUTION OF RESIDUALS
plot(WUE.aov,2)
```

# \*\*CROSSAIR DATA

### Load Libraries

```{r}
library(tidyverse) 
library(lattice) 
library(car) 
library(scales)
```

## DATA CLEANUP

```{r}
#load and select data
data.crosshair <- read.csv("GDE_crosshair.csv")
str(data.crosshair)

data.crosshair$Site <- as.factor(data.crosshair$Site)
data.crosshair$TotalDead <- (data.crosshair$DeadStanding + data.crosshair$DeadDown)
```

### Calculate Dead vs. Alive

```{r}
data.cross.dead <- data.crosshair %>% group_by (Site) %>% 
  summarise(sumDead = sum(TotalDead)) %>% ungroup()

data.cross.alive <- data.crosshair %>% group_by (Site) %>% 
  summarise(sumAlive = sum(LiveMesquite)) %>% ungroup()

data.cross.plot <- merge(
  data.cross.alive,
  data.cross.dead,
  by = "Site",
  all.x = TRUE) %>%
  pivot_longer(cols=c('sumAlive', 'sumDead'),names_to = "Status", values_to = "Count")
```

### Calculate Seedlings

```{r}
data.seed.plot <- data.crosshair %>%
  group_by(Site) %>%
  summarise(sumSeedling = sum(Seedings)) %>%
  ungroup()
```

## GRAPH

### Dead vs. Alive

```{r}
ggplot(data.cross.plot) + aes(x=Site, y=Count, fill=Status)+
  scale_fill_manual(values=c("darkgreen","chocolate4"),
                    labels=c('Alive', 'Dead')) +
  geom_col(alpha=1, position="dodge") + 
  scale_y_continuous(expand = c(0, 0),limits = c(0,25), 
                     breaks = pretty_breaks()) +
  scale_x_discrete(labels=c("1" = "Borrego Basin", "5" = "Clark Lake"))+
  theme_classic()+
  xlab("Site")+ylab("Count")+
  ggtitle("Mesquite Life Status at Each Survey Site")+
  theme(plot.title = element_text(hjust = 0.5))
```

### Seedlings

```{r}
ggplot(data.seed.plot) + aes(x=Site, y=sumSeedling)+
  scale_fill_manual()+
  geom_col(alpha=1, position="dodge") + 
  scale_y_continuous(expand = c(0, 0),limits = c(0,5), 
                     breaks = pretty_breaks()) +
  scale_x_discrete(labels=c("1" = "Borrego Basin", "5" = "Clark Lake"))+
  theme_classic()+
  xlab("Site")+ylab("Count")+
  ggtitle("Mesquite Seedling Counts at Each Survey Site")+
  theme(plot.title = element_text(hjust = 0.5))
```

## STATISTICAL ANALYSIS

### Dead

```{r}
#ANOVA
dead.aov <- aov(TotalDead ~ Site, data=data.crosshair)
summary(dead.aov)

#HOMOGENEITY OF VARIANCES
plot(dead.aov,1)
leveneTest(TotalDead ~ Site, data=data.crosshair)

#T-TEST
t.test(TotalDead~Site, data=data.crosshair)

#DISTRIBUTION OF RESIDUALS
plot(dead.aov,2)
```

### Alive

```{r}
#ANOVA
life.aov <- aov(LiveMesquite ~ Site, data=data.crosshair)
summary(life.aov)

#HOMOGENEITY OF VARIANCES
plot(life.aov,1)
leveneTest(LiveMesquite ~ Site, data=data.crosshair)

#T-TEST
t.test(LiveMesquite~Site, data=data.crosshair)

#DISTRIBUTION OF RESIDUALS
plot(life.aov,2)
```

### Seedlings

```{r}
#ANOVA
seed.aov <- aov(Seedings ~ Site, data=data.crosshair)
summary(seed.aov)

#HOMOGENEITY OF VARIANCES
plot(seed.aov,1)
leveneTest(Seedings ~ Site, data=data.crosshair)

#T-TEST
t.test(Seedings ~ Site, data=data.crosshair)

#DISTRIBUTION OF RESIDUALS
plot(seed.aov,2)
```

# \*\*Canopy Health

### Load Libraries

```{r}
library(tidyverse) 
library(lattice) 
library(car) 
library(scales)
library(GGally)
```

## DATA CLEANUP

```{r}
#load and select data
data.canopy <- read.csv("GDE_Canopy.csv")
str(data.canopy)

data.canopy$Site <- as.factor(data.canopy$Site) 
data.canopy$TreeID <- as.factor(data.canopy$Tree_ID)
```

### Check Normality

```{r}
histogram(data.canopy$Resprouting)
histogram(data.canopy$CrownSizeShape)
histogram(data.canopy$DeadBranches)
histogram(data.canopy$CrownFoliarDensity)
```

### Merge All Data

```{r}
#merge irga data and canopy health data by treeid
data.all <- merge(
  data.irga.indiv,
  data.canopy,
  by = "TreeID",
  all.x = TRUE) %>%
  subset(select=-Site.y) %>%
  rename(Site=Site.x)

```

## GRAPH

### Resprouting

```{r}
ggplot(data.canopy) + aes(x=Site, y=Resprouting, fill=Site)+
  scale_fill_manual(values=c("darkorange","darkgreen"),
                    labels=c('Borrego Basin', 'Clark Lake')) +
  geom_boxplot(aes(x=Site, y=Resprouting, fill=Site)) + 
  scale_x_discrete(labels=c("1" = "Borrego Basin", "5" = "Clark Lake"))+
  xlab("Site")+ylab("Resprouting") + 
  theme_classic()
```

### Crown Size Density

```{r}
ggplot(data.canopy) + aes(x=Site, y=CrownSizeShape, fill=Site)+
  scale_fill_manual(values=c("darkorange","darkgreen"),
                    labels=c('Borrego Basin', 'Clark Lake')) +
  geom_boxplot(aes(x=Site, y=CrownSizeShape, fill=Site)) + 
  scale_x_discrete(labels=c("1" = "Borrego Basin", "5" = "Clark Lake"))+
  xlab("Site")+ylab("Crown Size Shape") +
  theme_classic()
```

### Dead Branches

```{r}
ggplot(data.canopy) + aes(x=Site, y=DeadBranches, fill=Site)+
  scale_fill_manual(values=c("darkorange","darkgreen"),
                    labels=c('Borrego Basin', 'Clark Lake')) +
  geom_boxplot(aes(x=Site, y=DeadBranches, fill=Site)) + 
  scale_x_discrete(labels=c("1" = "Borrego Basin", "5" = "Clark Lake"))+
  xlab("Site")+ylab("Dead Branches") + 
  theme_classic()
```

### Crown Foliar Density

```{r}
ggplot(data.canopy) + aes(x=Site, y=CrownFoliarDensity, fill=Site)+
  scale_fill_manual(values=c("darkorange","darkgreen"),
                    labels=c('Borrego Basin', 'Clark Lake')) +
  geom_boxplot(aes(x=Site, y=CrownFoliarDensity, fill=Site)) + 
  scale_x_discrete(labels=c("1" = "Borrego Basin", "5" = "Clark Lake"))+
  xlab("Site")+ylab("CrownFoliar Density") + 
  theme_classic()
```

### All Data

```{r}
ggpairs(data.all, columns = c("Photo", "WUE","Cond", "Resprouting", "DeadBranches", "CrownFoliarDensity", "CrownSizeShape" ), title = "Bivariate analysis of BROMAD physiological traits", upper = list(continuous = wrap("cor",size = 3)),
        lower = list(continuous = wrap("smooth",
                                       alpha = 0.3,
                                       size = 0.1)),
        mapping = aes(color = Site.x)) +
  scale_fill_manual(values=c('darkorange','darkgreen')) +
  scale_colour_manual(values=c('darkorange','darkgreen'))
```

## STATISTICAL ANALYSIS

### Resprouting

```{r}
#ANOVA
resprout.aov <- aov(Resprouting ~ Site, data=data.canopy)
summary(resprout.aov)
#HOMOGENEITY OF VARIANCE
plot(resprout.aov,1)
leveneTest(Resprouting ~ Site, data=data.canopy)
#DISTRIBUTION OF RESIDUALS
plot(resprout.aov,2)
# EXTRACT RESIDUALS
resprout.aov_residuals <- residuals(object = resprout.aov )
#  SHAPIRO-WILK TEST
shapiro.test(x = resprout.aov_residuals )
```

### Crown Size Density

```{r}
#ANOVA
Crownsize.aov <- aov(Resprouting ~ Site, data=data.canopy)
summary(Crownsize.aov)
#HOMOGENEITY OF VARIANCE
plot(Crownsize.aov,1)
leveneTest(Resprouting ~ Site, data=data.canopy)
#DISTRIBUTION OF RESIDUALS
plot(Crownsize.aov,2)
# EXTRACT RESIDUALS
Crownsize.aov_residuals <- residuals(object = resprout.aov )
#  SHAPIRO-WILK TEST
shapiro.test(x = Crownsize.aov_residuals )

```

### Dead Branches

```{r}
#ANOVA
Deadbranches.aov <- aov(Resprouting ~ Site, data=data.canopy)
summary(Deadbranches.aov)
#HOMOGENEITY OF VARIANCE
plot(Deadbranches.aov,1)
leveneTest(Resprouting ~ Site, data=data.canopy)
#DISTRIBUTION OF RESIDUALS
plot(Deadbranches.aov,2)
# EXTRACT RESIDUALS
Deadbranches.aov_residuals <- residuals(object = Deadbranches.aov )
#  SHAPIRO-WILK TEST
shapiro.test(x = Deadbranches.aov_residuals )
```

### Crown Foliar Density

```{r}
#ANOVA
Density.aov <- aov(Resprouting ~ Site, data=data.canopy)
summary(Density.aov)
#HOMOGENEITY OF VARIANCE
plot(Density.aov,1)
leveneTest(Resprouting ~ Site, data=data.canopy)
#DISTRIBUTION OF RESIDUALS
plot(Density.aov,2)
# EXTRACT RESIDUALS
Density.aov_residuals <- residuals(object = Density.aov )
#  SHAPIRO-WILK TEST
shapiro.test(x = Density.aov_residuals )

```

FOCAL PROJECT FOR ADVANCED QUANT

```{r}
# Kelsey 
# maybe pca with data 

```
